---
title: "DK OIL"
author: 'c-r-loh'
date: "2025-02-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries and filtering of dataset as specified in instructions:
- Denmark
- Gas oil and diesel oil
- Gross inland deliveries - calculated
- Thousand tons (default)
90/10 Train/Test split
```{r, echo = FALSE, message=FALSE}
library(tidyverse)
library(forecast)
library(tseries)
library(FinTS)
data = read.csv("estat_nrg_cb_oilm_en.csv")
```

```{r}
oil = data %>% filter(
  geo == "Denmark" &
    siec == "Gas oil and diesel oil" &
    nrg_bal == "Gross inland deliveries - calculated"
) %>%
  dplyr::select(8, 9)

colnames(oil) = c("time", "obs")
oil$time = as.Date(paste(oil$time, "-01", sep =""))
train = oil[1:180,]; test = oil[181:nrow(oil),]
```

Initial Plot
- ACF shows a oscillatory pattern
  - Clear strong seasonal pattern with prominent spikes every 12 lags
- PACF shows large spike at lag 12, and gradually decays
  - Suggests seasonal dependency, gradual decay supports an AR structure
=> Indicates presence of seasonality, gradual decay suggesting seasonal AR
SARIMA will be modelled
```{r}

plot(oil$time, oil$obs, type = "l", ylab ="Value", xlab="Time")


acf(oil$obs, lag.max = 120, main = "", xaxt = "n")
axis(1, at = seq(0, length(oil$obs), by = 12))


pacf(oil$obs, lag.max = 120, main = "", xaxt = "n")
axis(1, at = seq(0, length(oil$obs), by = 12))

## for report
par(mfrow=c(1,2))
acf(oil$obs, lag.max = 60, main = "", xaxt = "n")
axis(1, at = seq(0, length(oil$obs), by = 12))


pacf(oil$obs, lag.max = 60, main = "", xaxt = "n")
axis(1, at = seq(0, length(oil$obs), by = 12))
```

Exploratory analyses
Decomposition
- Panel 2 indicates presence of seasonality
- Panel 3 indicates ambiguous trend

```{r}
#decomp on full
oil_ts_f = ts(oil$obs, frequency = 12)
decomp_f = stl(oil_ts_f, s.window = "periodic")
plot(decomp_f)


#decomp on training
oil_ts = ts(train$obs, frequency = 12)
decomp = stl(oil_ts, s.window = "periodic")
plot(decomp)
resid = decomp$time.series[, "remainder"]

```

Manually checking for trend
- Expecting an eventual downward trend post-2013 due to change in regulation


```{r}
#estimating trend component
decomp = decompose(oil_ts, type = "additive")
rem_season = oil_ts - decomp$seasonal
autoplot(rem_season)
adf.test(rem_season) #stationary since pval<0.05, no differencing required..?

#modelling trend using interaction since change in behaviour should be expected to be gradual
train$policy = ifelse(train$time >= as.Date("2013-01-01"), 1, 0)
time_index = 1:length(rem_season)
mod_policy = lm(rem_season~time_index *policy, data = train)
summary(mod_policy)
checkresiduals(mod_policy)
#acf results indicate some autocorrelation (lags ~8)
#residual histogram close to normally distributed
#can be improved by including lagged predictors

ggplot(train, aes(x = time, y = obs)) +
  geom_line() +
  geom_line(aes(y = predict(mod_policy)), color = "red") +
  geom_vline(xintercept = as.Date("2013-01-01"), linetype="dashed", color="blue") +
  ggtitle("Trend Model with Policy Intervention")


future_time_index = seq(max(time_index) + 1, max(time_index) + nrow(test))
test$time_index = future_time_index
test$policy = rep(1, nrow(test))
future_trend = predict(mod_policy, newdata = test)

test$future_trend = predict(mod_policy, newdata = test)

# merge train and test data
train$future_trend = NA  
forecast_df = bind_rows(train, test)  # Combine both datasets

# visualizing forecast
ggplot(forecast_df, aes(x = time, y = obs)) +
  geom_line() +
  geom_line(aes(y = future_trend), color = "red") +
  geom_vline(xintercept = as.Date("2013-01-01"), linetype = "dashed", color = "blue") +
  ggtitle("Forecast Adjusted for Legislation Impact")


```

Modelling seasonality with dummy variables
```{r, DV}
train$month = factor(format(train$time, "%m"))
mod_season_dummy = lm(obs ~ month, data = train)
summary(mod_season_dummy)
test$month = factor(format(test$time, "%m"), levels = levels(train$month))
test$future_season = predict(mod_season_dummy, newdata = test)

ggplot() +
  geom_line(data = train, aes(x = time, y = obs), color = "black") +
  geom_line(data = test, aes(x=time, y=obs), color = "black") +
  geom_line(data = test, aes(x = time, y = future_season), color="red", linetype = "dashed") +
  geom_line(data = train, aes(x = time, y=predict(mod_season_dummy, newdata = train)), color = "red") +
  theme_minimal() +
  ggtitle("Seasonal Model with Forecast")

#checking residuals
train$resid = residuals(mod_season_dummy)

ggAcf(train$resid) + ggtitle("ACF") #lag 9 term
ggPacf(train$resid) + ggtitle("PACF") 

train$lag9 = lag(train$obs, 9)
test$lag9 = lag(test$obs, 9)
#adding lagged to seasonal
mod_season_dummy_lag = lm(obs ~ month + lag9, data = train, na.action = na.omit)

ggAcf(residuals(mod_season_dummy_lag)) #lag 8

train$lag8 = lag(train$obs, 8)
test$lag8 = lag(test$obs, 8)
test$lag9[1:9] = train$obs[(nrow(train)-9): nrow(train)]
test$lag8[1:8] = train$obs[(nrow(train)-8): nrow(train)]

mod_season_dummy_lag_2 = lm(obs ~ month + lag9 + lag8, data = train, na.action = na.omit)

ggAcf(residuals(mod_season_dummy_lag_2))
adf.test(residuals(mod_season_dummy_lag_2))

#for predicting 2025 values
future_test = data.frame(
  time = seq(max(test$time), by = "month", length.out=5)[-1],
  month = factor(format((seq(max(test$time), by = "month", length.out=5)[-1]), "%m"), levels = levels(train$month))
)
future_test$lag9 = c(tail(test$obs,9)[1:4])
future_test$lag8 = c(tail(test$obs,8)[1:4])

future_test$pred = predict(mod_season_dummy_lag_2, newdata = future_test)


ggplot() +
  geom_line(data = train, aes(x = time, y = obs), color = "gray") +
  geom_line(data = train, aes(x = time, y=predict(mod_season_dummy_lag_2, newdata = train)), color = "red") + 
  geom_line(data = test, aes(x=time, y=obs), color = "gray") +
  geom_line(data = test, aes(x = time, y = predict(mod_season_dummy_lag_2, newdata = test)), color="red", linetype = "dashed") +
  geom_line(data = future_test, aes(x=time, y=pred), color = "blue", linetype = "dashed") +
  theme_minimal() +
  ggtitle("Seasonal Model with two lags")

summary(mod_season_dummy_lag_2)
AIC(mod_season_dummy_lag_2)
resid_dv = residuals((mod_season_dummy_lag_2))
ggAcf(resid_dv, lag.max = 120)
ggPacf(resid_dv, lag.max = 120)
adf.test(resid_dv)
```


auto.arima to ballpark
```{r}

mod_ballpark = auto.arima(rem_season)
summary(mod_ballpark)

best_arima = auto.arima(oil_ts)
summary(best_arima)

best_sarima = auto.arima(oil_ts, seasonal = TRUE)
summary(best_sarima)
```

Box-Jenkins Breakdown
1. Model Identification
- Stationarity 
- Order Selection

2. Model Estimation
- Differencing if required
- Recheck ACF/PACF

3. Fitting & Diagnostics
- Ljung-Box, ACF and QQ on residuals

4. Forecasting & Validation

1. Model Identification
```{r, modelIdentification}
oil_ts = ts(train$obs, start = c(2008, 01),frequency = 12)
autoplot(oil_ts) + ggtitle("Training set")
ggAcf(oil_ts, lag.max = 120) + ggtitle("ACF")
ggPacf(oil_ts, lag.max = 120) + ggtitle("PACF")
adf.test(oil$obs, k = 5)
adf.test(oil_ts, k = 5)
```
ACF & PACF graphs indicate strong seasonality that needs to be modelled
- Seasonal terms in ACF has sinusoidal pattern - spikes at 12, 24 that gradually decays 
- Annual seasonality
- Short-term correlations
=> Seasonal MA may be required

- PACF spike at 12 and decays over time 
=> Seasonal AR term may be required

adf.test() indicates lag terms ~5
=> No differencing required since stationary.

2. Model Estimation
```{r, modelEst}
results = list()
#range of values to check
#p = 0:1; q = 0:1; 
P_vals = 0:2; Q_vals = 0:2

#for (p in p) {
#  for (q in q) {
for (P in P_vals) {
  for (Q in Q_vals) {
    model = paste0("SARIMA(0,0,0)(", P, ",0,", Q, ")[12]")
    mod = tryCatch({
      arima(
        oil_ts,
        order = c(0, 0, 0),
        seasonal = list(order = c(P, 0, Q), period = 12)
      )
    }, error = function(e)
      return(NULL))
    
    if (is.null(mod)) {
      cat("Error", model, "skip")
      next
    }
    
    results[[model]] = list(
      model = mod,
      train_pred = fitted(mod),
      test_pred = forecast(mod, h = nrow(test))$mean,
      aic = mod$aic
    )
  }
}
#  }
#}

aic_vals = sapply(results, function(x) x$aic)
aic_vals

#selection between sarima(0,0,0)(1,0,0)[12] and sarima(0,0,0)(2,0,1)[12] based on aic, going with the simpler model since aic difference minute 

chosen_sarima = results[["SARIMA(0,0,0)(1,0,0)[12]"]][["model"]]
summary(chosen_sarima)

alt_sarima = results[["SARIMA(0,0,0)(2,0,1)[12]"]][["model"]]
summary(alt_sarima)
#checking residuals
resid_sarima = residuals(chosen_sarima)
ggAcf(resid_sarima,lag.max = 120)
ggPacf(resid_sarima, lag.max = 120)
#no significant spikes
```

3. Fitting & Diagnostics
```{r}
ljungbox_12 = Box.test(resid_sarima, lag = 12, type = "Ljung-Box", fitdf = 1)

ljungbox_24 = Box.test(resid_sarima, lag = 24, type = "Ljung-Box", fitdf = 1)

ljungbox_12; ljungbox_24
#ljung-box indicates residuals are white noise (pval > 0.05)
```
4. Forecasting & Validations

```{r}
forecast_sarima = forecast(chosen_sarima, h = nrow(test), level = c(90,95))
test$pred_sarima = forecast_sarima$mean
test$lower_95 = forecast_sarima$lower[,2]
test$upper_95 = forecast_sarima$upper[,2]
train$pred_sarima = fitted(chosen_sarima)

ggplot() +
  geom_line(data = train, aes(x = time, y = obs), color = "gray") +
  geom_line(data = train, aes(x = time, y = pred_sarima), color = "red") +
  geom_line(data = test, aes(x= time, y=obs), color = "gray") +
  geom_line( data = test, aes(x=time, y=pred_sarima), color = "red", linetype = "dashed") + 
  geom_ribbon(data = test, aes(x=time, ymin = lower_95, ymax = upper_95), fill = "red", alpha = 0.2) +
  theme_minimal() +
  ggtitle("SARIMA forecast with 95% CI")

accuracy(forecast_sarima, test$obs) #look at percentage


#forecasting dec'24 - mar'25
forecast_extend = nrow(test) + 4
forecast_sarima_ext = forecast(chosen_sarima, h = forecast_extend, level = c(90,95))
forecast_dates = seq(max(test$time), by = "month", length.out = 5)[-1]

#adjusting dataframes for presentation
ext_test = data.frame(
  time = forecast_dates,
  pred_sarima = forecast_sarima_ext$mean[(nrow(test)+1):forecast_extend],
  lower_95 = forecast_sarima_ext$lower[(nrow(test)+1): forecast_extend, 2],
  upper_95 = forecast_sarima_ext$upper[(nrow(test)+1): forecast_extend, 2]
)
observed_data = oil %>%
  filter(time >= "2022-01-01")

connect = train %>%
  filter(time ==max(time)) %>%
  mutate(lower_95 = NA, upper_95= NA)
forecast_combined = bind_rows(connect, test, ext_test)

ggplot() +
  geom_line(data = observed_data, aes(x=time, y=obs), color = "gray") +
  geom_line(data = train[train$time >= as.Date("2022-01-01"),], aes(x=time, y=pred_sarima), color="red") +
  geom_line(data = forecast_combined, aes(x=time, y=pred_sarima), color="red") +
  geom_ribbon(data = forecast_combined, aes(x=time, ymin=lower_95, ymax=upper_95), fill="red", alpha=0.1) +
  geom_line(data=forecast_combined[forecast_combined$time>=as.Date("2024-12-01"),], aes(x=time, y=pred_sarima), color="blue") +
  geom_ribbon(data = forecast_combined[forecast_combined$time >= as.Date("2024-12-01"), ], aes(x=time, ymin=lower_95, ymax=upper_95), fill="blue", alpha=0.1) +
  theme_minimal()+
  ggtitle("Extended SARIMA Forecast with 95% CI")
##heteroscedasticity
plot(train$pred_sarima, residuals(chosen_sarima))
qqnorm(residuals(chosen_sarima)) #appear normal, few outliers
qqline(residuals(chosen_sarima))

white.test(residuals(chosen_sarima)) #at 0.05 no evidence of heteroscedasticity -> CI estimates of forecast valid, GLS not required

```


```{r}

##
ggplot() +
  geom_line(data = train[train$time >= as.Date("2022-01-01"),], aes(x = time, y = obs), color = "gray") +
  geom_line(data = train[train$time >= as.Date("2022-01-01"),], aes(x = time, y = pred_sarima), color = "red") +
  geom_line(data = forecast_combined, aes(x= time, y=obs), color = "gray") +
  geom_line(data = forecast_combined, aes(x=time, y=pred_sarima), color = "red") + 
  geom_ribbon(data = forecast_combined, aes(x=time, ymin = lower_95, ymax = upper_95), fill = "red", alpha = 0.2) +
  geom_line(data = forecast_combined[forecast_combined$time >= as.Date("2024-12-01"),], aes(x=time, y=pred_sarima), color = "blue") +
  geom_ribbon(data = forecast_combined[forecast_combined$time >= as.Date("2024-12-01"),], aes(x = time, ymin = ts(lower_95), ymax = upper_95), fill = "blue", alpha = 0.2) +
  theme_minimal() +
  ggtitle("Extended SARIMA forecast with 95% CI")

##recheck forecast
```

```{r, test}
ggplot() +
  # observed data (gray)
  geom_line(data = observed_data, aes(x = time, y = obs), color = "gray") +

  # forecasted predictions (red)
  geom_line(data = forecast_combined, aes(x = time, y = pred_sarima), color = "red") +

  # unified forecast confidence interval (single ribbon)
  geom_ribbon(data = forecast_combined, aes(x = time, ymin = lower_95, ymax = upper_95), 
              fill = "gray", alpha = 0.2) +   # Match second image style

  theme_minimal() +
  ggtitle("Extended SARIMA Forecast with Unified 95% CI")
```

Tables and figures
```{r}
library(MLmetrics)
library(knitr)
library(kableExtra)
#aic and mape comparison
aic_values = c(
  round(AIC(mod_season_dummy_lag_2),2),
  round(AIC(chosen_sarima),2),
  round(AIC(alt_sarima),2)
)
#rmse mae mape
fcv = as.numeric(forecast(chosen_sarima, h=nrow(test))$mean)

rmse_sar = round(sqrt(mean((fcv - test$obs)^2)),3)
mae_sar = round(mean(abs(fcv - test$obs)),3)
rmse_sar; mae_sar

mape_values = c(
  round(MAPE(predict(mod_season_dummy_lag_2, newdata = test), test$obs),2),
  round(MAPE(forecast(chosen_sarima, h =nrow(test))$mean, test$obs),2),
  round(MAPE(forecast(alt_sarima, h = nrow(test))$mean, test$obs),2)
)

model_names = c("DV+Lagged", "SAR(1)", "SARIMA(0,0,0)(2,0,1)[12]")

comparison_table = data.frame(
  Model = model_names,
  AIC = aic_values,
  MAPE = mape_values
)
comparison_table = comparison_table %>%
  arrange(AIC)

kable(comparison_table)

#forecast graph


ggplot() +
  #observed
  geom_line(data = train[train$time >= as.Date("2022-01-01"),], aes(x = time, y = obs, color = "Observed Data")) +
    geom_line(data = forecast_combined, aes(x= time, y=obs, color = "Observed Data")) +
  
  #fitted
  geom_line(data = train[train$time >= as.Date("2022-01-01"),], aes(x = time, y = pred_sarima, color = "Fitted Data")) +
  geom_line(data = forecast_combined, aes(x=time, y=pred_sarima, color = "Fitted Data")) + 
  geom_ribbon(data = forecast_combined, aes(x=time, ymin = lower_95, ymax = upper_95), fill = "red", alpha = 0.2) +
  
  #forecast
  geom_line(data = forecast_combined[forecast_combined$time >= as.Date("2024-12-01"),], aes(x=time, y=pred_sarima, color = "Forecast")) +
  geom_ribbon(data = forecast_combined[forecast_combined$time >= as.Date("2024-12-01"),], aes(x = time, ymin = ts(lower_95), ymax = upper_95), fill = "blue", alpha = 0.2) +
  ggtitle("Extended SAR(1) Forecast with 95% CI") +
  xlab("Time") +
  ylab("Oil Units") +
  scale_color_manual(
    name = "Legend",
    values = c("Observed Data" = "gray",
               "Fitted Data" = "red",
               "Forecast" = "blue")) +
  theme(legend.position = c(0.1,0.93),
        legend.key = element_rect(fill = "white"))


#forecast table
ext_test_kable = ext_test %>%
  mutate(CI = paste0("(", round(lower_95,2), ", ", round(upper_95,2), ")")) %>% 
  select(time, pred_sarima, CI)
colnames(ext_test_kable) = c("Date", "Forecasted Value", "95% CI")

fv_kable = ext_test_kable %>%
  mutate(`Forecasted Value` = round(`Forecasted Value`, 2))

#model assumptions and test results
#stationarity, adf pval <0.05
adf.test(residuals(chosen_sarima))
#=> stationary
#on residuals
checkresiduals(chosen_sarima) 
#residuals approximately normal, looks centered about 0

Box.test(residuals(chosen_sarima), lag = 12, type = "Ljung-Box") #no serial ac, pval >0.05
white.test(residuals(chosen_sarima))
#homoscedastic, pval >0.05


#=> residuals white noise, approximately 0 mean with constant variance

summary(chosen_sarima)

tsdiag(chosen_sarima)
```