---
title: "MATH453 Coursework Final"
author: '36775864'
date: "2025-03-25"
output: pdf_document
---

```{r, echo = F, message= F}
library(lme4)
library(tidyverse)
library(nlme)
load("Data1.RData")
library(viridis)
library(lmtest)
library(reshape2)
library(statmod)
library(DHARMa)
#library(lmerTest)
load("Data2.RData")
```

## Problem 1
- Count data
- Overdispersed
```{r}
var(Data1$Score)/mean(Data1$Score)
```


```{r, Q1FITCHECKS}
mod.nb = glmer.nb(Score ~ Age * Gender + (1 + Age|ReaserchGroupID), data = Data1)
mod.nb1 = glmer.nb(Score ~ Age * Gender + (1|ReaserchGroupID), data = Data1)

mod.bin = glmer(cbind(Score,(100-Data1$Score)) ~ Age * Gender + (1 + Age|ReaserchGroupID), family = "binomial", data = Data1)

anova(mod.nb, mod.nb1)
```

```{r, Q1SIM}
sim = simulate(mod.nb, nsim = 10000, seed = 11)
#Score of Age 40-59 lower than 20-39
old_rows = which(Data1$Age =="40-59")
young_rows = which(Data1$Age == "20-39")

sim_old  = sapply(sim, function(x) mean(x[old_rows]))
sim_young = sapply(sim, function(x) mean(x[young_rows]))
#require mean(x[.]) to reduce to one summary statistic per group per sim
hist((sim_old-sim_young), density = TRUE)
prob_q1 = mean(sim_old < sim_young)
nb_means = c(mean(sim_old), mean(sim_young))

theta = getME(mod.nb, "glmer.nb.theta")
```

```{r, Q1SIMVISUALCHECK}
sim_mean = rowMeans(sim)
df_combined_hist = data.frame(
  value = c(Data1$Score, sim_mean),
  source = rep(c("Observed", "Simulated"), c(length(Data1$Score), length(sim_mean)))
)

ggplot(df_combined_hist, aes(x =value, fill = source)) +
  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 5) +
scale_fill_manual(values = c("Observed" = "black", "Simulated" = "skyblue")) +
  theme_minimal() +
  labs(x = "Value", y = "Count", title = "Histogram: Observed vs Simulated Mean")

```





##Problem 2
-RID for problem 2 irrelevant
```{r}

pairs(Data2)

ggplot(data=Data2,aes(y=logScore,x=time, group= PatientID )) +
  geom_line()+
  geom_point()+
  facet_grid(. ~ Gender)

hist(Data2$logScore, breaks = 10)
```
looks relatively linear
TRY LINEAR

Time fixed effect
Random intercept per PatientID

```{r, Q2COVARIATES}

mod.lm1 = lmer(logScore ~ Age + Gender + time + (1 | PatientID) + (0 + time | PatientID), data = Data2)
mod.lm2 = lmer(logScore ~ Age + Gender * time + (1 + time | PatientID), data = Data2)
mod.lm3 = lmer(logScore ~ Age + Gender + time + (1 + time | PatientID), data = Data2)
mod.lm4 = lmer(logScore ~ Age + Gender + Gender * time + (1 + time | PatientID), data = Data2)

#comparing interaction terms


AIC(mod.lm1, mod.lm2, mod.lm3, mod.lm4)

```

```{r}
summary(mod.lm3)
```
-Quantify growth curve
All else constant, the expected improvement of logScore improves by 0.015 log units per month.

-P(logScore > log(90) | t = 18)
That is, an patient with unspecified characteristics has a 34% chance of having logScore > log(90) 18 months post implementation.
```{r, Q2SIM}
sim_q2 = simulate(mod.lm3, nsim = 10000, seed = 11)
t18_rows = which(Data2$time == 18)
#t0_rows = which(Data2$time == 0)

#sim_t0 = sapply(sim_q2, function(x) (x[t0_rows]))
sim_t18  = sapply(sim_q2, function(x) (x[t18_rows]))
prob_q2 = mean(sim_t18 > log(90))
#prob_compare = mean(sim_t0 > log(90))

prob_q2; #prob_compare
```

```{r}
check1 = Data2 %>%
  filter(time == 18, logScore > log(90))
check2 = Data2 %>%
  filter(time == 18)
nrow(check1)/nrow(check2)
```

- Compare M/F growth curves
No evidence that gender have different rates of change; they will have the same curve albeit different intercept as seen frmo mod.lm3.
```{r, Q2ANOVA}
#mod.lm3 = lmer(logScore ~ Age + Gender + time + (1 + time | PatientID), data = Data2)
mod.lm3.1 = lmer(logScore ~ Age + Gender * time + (1 + time | PatientID), data = Data2)
mod.lm3.2 = lmer(logScore ~ Age * Gender + time + (1 + time| PatientID), data = Data2)
anova(mod.lm3, mod.lm3.1)
anova(mod.lm3, mod.lm3.2)

```

```{r, Q2RESID}
qqnorm(resid(mod.lm3), ylab = "Residuals", main = ""); qqline(resid(mod.lm3))

plot(resid(mod.lm3),
     ylab = "Residuals")
```


Prettied plots for report
```{r}
score_range = seq(0,100,1)
size = 100
mu = mean(Data1$Score)
sigmasq = var(Data1$Score)
r = (mu**2)/(sigmasq - mu)
p = mu/sigmasq
nb = dnbinom(seq(0,100,1), size = r, prob = p)
nb_df = data.frame(seq(0,100,1), nb)
bin = dbinom(seq(0,100,1), size = size, prob = mu/size)
bin_df = data.frame(seq(0,100,1), bin)
norm = dnorm(seq(0,100,1), mean = mu, sd = sqrt(sigmasq))
norm_df = data.frame(seq(0,100,1), norm)

#final plot
ggplot(Data1, aes(x = Score)) +
  geom_histogram(aes(y = after_stat(density)), breaks = seq(0,100, by=5), alpha = 0.5) +
  geom_line(data = nb_df, aes(x = score_range, y = nb), color = "red", size = 1) +
  geom_line(data = bin_df, aes(x = score_range, y = bin), color = "blue", size = 1, lty = 8) +
  geom_line(data = norm_df, aes(x = score_range, y = norm), color = "purple", size = 1, lty = 8) +
  theme_minimal()

#prettied up
q1plot = ggplot(Data1, aes(x = Score)) +
  geom_histogram(
    aes(y = after_stat(density), fill = "Observed Data"),
    breaks = seq(0, 100, by = 5),
    alpha = 0.5,
    color = "black"
  ) +
  geom_line(
    data = nb_df,
    aes(
      x = score_range,
      y = nb,
      color = "Negative Binomial",
      linetype = "Negative Binomial"
    ),
    size = 1
  ) +
  geom_line(
    data = bin_df,
    aes(
      x = score_range,
      y = bin,
      color = "Binomial",
      linetype = "Binomial"
    ),
    size = 1
  ) +
  geom_line(
    data = norm_df,
    aes(
      x = score_range,
      y = norm,
      color = "Gaussian",
      linetype = "Gaussian"
    ),
    size = 1
  ) +
  
  scale_fill_manual(name = NULL,
                    values = c("Observed Data" = "grey80")) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Negative Binomial" = "firebrick",
      "Binomial" = "steelblue",
      "Gaussian" = "darkorchid"
    )
  ) +
  scale_linetype_manual(
    name = NULL,
    values = c(
      "Negative Binomial" = "solid",
      "Binomial" = "dotted",
      "Gaussian" = "dotted"
    )
  ) +
  ggtitle("Histogram of Score with theoretical PDFs") +
theme_minimal(base_size = 14) + 
theme(
  legend.position = c(0.05, 0.95),
  legend.justification = c(0, 1),
  legend.background = element_blank(),
  legend.key = element_blank(),
  legend.text = element_text(size = 12),        
  axis.title = element_text(size = 14, face = "bold"),  
  axis.text = element_text(size = 12),           
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  
  panel.grid.major = element_line(color = "grey80", linewidth = 0.3), 
  panel.grid.minor = element_blank()
)
q1plot


mu2 = mean(Data2$logScore)
sigmasq2 = var(Data2$logScore)
norm2 = dnorm(seq(-11,17,1), mean = mu2, sd = sqrt(sigmasq2))
norm_df2 = data.frame(seq(-11,17,1), norm2)

ggplot(Data2, aes(x = logScore)) +
  geom_histogram(aes(y = after_stat(density)),
                 breaks = seq(-11, 17, by = 1),
                 alpha = 0.5) +
  geom_line(
    data = norm_df2,
    aes(x = seq(-11, 17, 1), y = norm2),
    color = "firebrick",
    size = 1
  )

q2plot = ggplot(Data2, aes(x = logScore)) +
  geom_histogram(
    aes(y = after_stat(density), fill = "Observed Data"),
    breaks = seq(-11, 17, by = 1),
    alpha = 0.5,
    color = "black"
  ) +
  geom_line(data = norm_df2,
            aes(
              x = seq(-11, 17, 1),
              y = norm2,
              color = "Gaussian PDF"
            ),
            size = 1) +
  
  scale_fill_manual(name = NULL, values = c("Observed Data" = "grey70")) +
  scale_color_manual(name = NULL, values = c("Gaussian PDF" = "firebrick")) +
  
  ggtitle("Histogram of logScore with theoretical PDF") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position= c(0.05, 0.95),
    legend.justification = c(0, 1),
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "grey80", linewidth = 0.3),
    panel.grid.minor = element_blank()
  ) +
  guides(
    fill = guide_legend(override.aes = list(fill = "grey60")),
    color = guide_legend(override.aes = list(linewidth = 1.5))
  )

q1plot
q2plot
```